Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-07 paper/main.tex
Merge branch

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{todonotes}
\usepackage{bm}
\usepackage{subcaption}
% \usepackage{fancyvrb}
\usepackage[ruled]{algorithm2e}

@@

\section{Using Measure of Association for Ranking and DAG Construction}
\label{sec:modification}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{todonotes}
\usepackage{bm}
\usepackage{subcaption}
\usepackage[linesnumbered, ruled]{algorithm2e}

@@

\section{DAG Structure Learning using Ancestral Oracles}
\label{sec:modification}

Our approach to causal structure learning combines domain knowledge with data-driven
insights in a manner that is based on the following base assumptions: 
(1) Domain experts are good at determining causal directions between variables if 
there is a clear causal direction between them. (2) Domain experts are less good
at identifying cases where there is no causal relationship between the variables,
since \emph{potential} causal relationships can often be argued for anyway.
(3) Domain experts are even less good at distinguishing direct from indirect 
effects, since the question whether a direct effect exists between two variables
depends on all other variables present in the graph.

We therefore consider the following two models of a domain experts: Given two variables
\emph{X} and \emph{Y} that are part of an unknown causal DAG $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$

whereas a \emph{weak ancestral oracle} is defined as:
$$
\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{rand}(X \gets Y, X \to Y) & \textrm{otherwise} \\
\end{cases}
$$

That is, our ancestral oracles never distinguish whether the relationships between variables is
direct or indirect; and the weak ancestral oracle is even unable to detect cases where there is no
cause-effect relationship between two variables. Using such oracles for DAG construction therefore
inherently risks that superfluous edges are created. In our approach, such superfluous edges will
be detected and removed by utilizing the data.

To model how our learning procedure interacts with the data, we
assume that we have access to a second oracle $\mathcal{D}_G$ than can answer
d-separation queries with respect to the unknown target graph: 
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X \ci_G Y \mid \mathbf{Z}$. 
These are the standard oracles considered in constraint-based structure learning algorithms,
such as the PC algorithm \cite{Kalisch2012}. Based on this oracle, we now define two procedures
that iteratively change a current DAG structure.

Our structure learning algorithm is based on two core procedures. \textsc{Expand} 
(Algorithm~\ref{algo:expand}) uses conditional independence information to search for
missing edges in the graph and then uses domain knowledge to orient them. 
\textsc{Prune} (Algorithm~\ref{algo:prune}) uses conditional independence information
to remove superfluous edges from the graph.

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B=\emptyset$,$k=\infty$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
            $L \gets L \cup \mathcal{A}_G(X, Y)$ \;
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data an domain knowledge}
\label{algo:expand}
\end{algorithm}

The function $\textsc{Expand}$ takes an initial list of edges and searches for 
any vertex pairs that are not connected but where the d-separation oracle 
indicates  that there is still a residual association not explained by the other variables.
An optional parameter $B$, the empty set by default, allows to specify a ``black list''
of vertex pairs that must not be connected, and another optional parameter $k$
specifies the maximum amount of edges to be added by this procedure.

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a strong ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G)=G^+$.
\end{proposition}

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a weak ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G)\supseteq G^+$.
\end{proposition}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined

\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Prune}($V$,$E$,$\mathcal{D}_G$)}{
    let $L \gets \{\}$\;
    \ForEach{$X \to Y \in E$}{
        remove $X \to Y$ from $E$\;
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 1$}{
            $L \gets L \cup  \{X \to Y\}$ \;
        }
    }
    remove all edges in $L$ from $G$\;
    \Return{$(V,E)$}\;
}
\caption{Pruning superfluous edges}
\label{algo:prune}
\end{algorithm}

\begin{proposition}
Consider two DAGs $G=(V,E)$ and $G'=(V,E')$ where $E \subseteq E'$. 
Then $\textsc{Prune}(G',\mathcal{D}_G)=G$.
\label{prop:prune}
\end{proposition}

By combining the previous three propositions, we immediately obtain the following:

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a strong or weak 
ancestral oracle $\mathcal{A}_G$, 
$\textsc{Prune}(\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_G),\mathcal{D}_G)=G.$
\end{theorem}

Since we require expert knowledge only in the $\textsc{Expand}$ operation, we may try
to be more economical by asking fewer questions at a time and interleaving expansion and
pruning steps. This leads us to the following, more incremental DAG construction algorithm.

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwProg{Fn}{Function}{:}{}

\Fn{{\sc ExpertInLoop}($V,\mathcal{D}_G,\mathcal{A}_G$)}{
$E_p \gets \emptyset$ \tcc{Current edges} 
$B \gets \emptyset$ \tcc{Edges that were pruned} 
\Repeat{ $E=E_p$ }{
	$E \gets E_p$ \;
	$(V,E) \gets \textsc{Expand}(V,E,\mathcal{D}_G,\mathcal{A}_G,B,1)$ \;
	$(V,E_p) \gets \textsc{Prune}(V,E,\mathcal{D}_G)$ \;
	$B \gets B \cup \{ E_p \setminus E \} $
}
\Return{(V,E)}
}

\caption{Iterative structure learning with expert in the loop}
\label{algo:expert}
\end{algorithm}

\begin{theorem}
Let $G=(V,^*E)$ be a DAG, $\mathcal{D}_G$ a d-separation oracle for $G$, and 
$\mathcal{A}_G$ a strong or weak ancestral oracle for $G$. Then 
$\textsc{ExpertInLoop}(V,\mathcal{D}_G,\mathcal{A}_G)=G$.
\end{theorem}

\begin{proof}
The loop in Algorithm~\ref{algo:expert} terminates if and only if $E=E^*$. If the loop does not 
terminate, then a new edge has been added in line 6, or one or more edges were pruned in line 7.
Every edge can be added at 
most once and pruned at most once. Therefore, the algorithm always terminates
after at most $|V|(|V|-1)+1$ iterations of the loop. For every edge $e=X\to Y $ in the skeleton of $G$, $\mathcal{D}_G(X,Y,\mathbf{Z})=0$ irrespective of $\mathbf{Z}$. Therefore, $e$ must be added to $E$ in some iteration, and can never be pruned afterwards. Therefore, after some iteration, $(V,E)$ must be a supergraph of $G$ after executing line 6, and this will be pruned to the real graph $G$ in line 7 (Proposition~\ref{prop:prune}). In the next iteration, no further changes are made, and the loop terminates.
\end{proof}
